Variance reduction for MCMC methods via martingale representations

### 2020-04-10 aap/Revision/vr-mcmc-revision-new-new.tex
update

Y_p, & \texttt{ with probability } \alpha(x,Y_p),

@@

x, &  \texttt{ with probability } 1-\alpha(x,Y_p),

@@

\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.

@@

On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)

@@

&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].

@@

for any integrable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator

@@

The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.

@@

for some $l_0$ with $l_0+q\leq N+n,$

@@

\widehat{A}_{q,k}(y)=

@@

\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].

@@

@@

\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2

@@

 as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover,

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$. At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
\section{Analysis of variance reduced ULA}

@@

H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},

@@

\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.

@@

M_{K,n}^N(f) & =\sum_{ 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})

@@

A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).

@@

\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that

>>>>

Y_p, & \texttt{ with probability } \alpha_p(x,Y_p),

@@

x, &  \texttt{ with probability } 1-\alpha_p(x,Y_p),

@@

\alpha_p(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.

@@

%On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)

@@

&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l}).

@@

for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator

@@

The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way to estimate them is to use nonparametric regression.

@@

for $k=1,\ldots,K,$ and some $l_0$ with $l_0+q\leq N+n,$

@@

\widehat{A}_{q,k}(y):=

@@

\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].

@@

\label{eq:PVarA}

@@

\bar A_{q,k}:=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2

@@

 as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover, it follows from \eqref{eq:PVarA} that

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n$ and $k=1,\ldots,K$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$\section{Analysis of variance reduced ULA}

@@

H_k(x):=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},

@@

\mathbf{H}_\mathbf{k}(\mathbf{x}):=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.

@@

M_{K,n}^N(f) & :=\frac{1}{n}\sum_{ 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})

@@

A_{q,\mathbf{k}}(y):=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).

@@

\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz continuous, that is, there exists $L < \infty$ such that
